Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-26 appendix.tex
regular update, code updates
editing numerical results, redid mmsn application for 50au and corrected espilon to include mu dependence

\section{Fiducial model for a protoplanetary disk}\label{mmsn}
For results application in \S\ref{application} we use the disk model
of \cite{chiang10}. This disk model orbits a Solar-mass star and 
has a surface density distribution
\begin{align}\label{mmsn_sigma}
  \Sigma = 2200
  \hat{\Sigma}\left(\frac{r}{\mathrm{AU}}\right)^{-3/2}\mathrm{g}\,\mathrm{cm}^{-2},  
\end{align}
a mid-plane density distribution
\begin{align}
  \rho_0 = 2.7\times10^{-9}
  \hat{\Sigma}\left(\frac{r}{\mathrm{AU}}\right)^{-39/14}\mathrm{g}\,\mathrm{cm}^{-3},  
\end{align}
which implies $p=-39/14$, 
a temperature profile
\begin{align}\label{mmsn_temp}
  T = 120\hat{T}\left(\frac{r}{\mathrm{AU}}\right)^{-3/7} \mathrm{K},
\end{align}
which implies $q=-3/7$, 
and the fiducial disk aspect-ratio 
\begin{align}
  \epsilon = 0.022\left(\frac{r}{\mathrm{AU}}\right)^{2/7}.
\end{align}
In the above expressions, $\hat{\Sigma}$ and
$\hat{T}$ are dimensionless coefficients used to scale the model
relative to the MMSN. Their nomial values are unity. In addition, we
use the opacity model {\bf(reference?)}
 \begin{align}
   \kappa_d &= 2 \hat{\kappa}_d \left(\frac{T}{100\mathrm{K}}\right)^2
   \mathrm{cm}^2\,\mathrm{g}
    =
   2.88\hat{\kappa}_d\hat{T}^2\left(\frac{r}{\mathrm{AU}}\right)^{-6/7}\mathrm{cm}^2\,\mathrm{g}^{-1},   
 \end{align}
where the second equality follows from the model temperature profile
above, and $\hat{\kappa}_d$ has similar meaning as $\hat{\Sigma}$ and
$\hat{T}$.

>>>>

\section{Fiducial model for a protoplanetary disk}\label{mmsn}
For results application in \S\ref{application} we use the disk model
of \cite{chiang10}. This disk model orbits a Solar-mass star and 
has a surface density distribution
\begin{align}\label{mmsn_sigma}
  \Sigma = 2200
  \hat{\Sigma}\left(\frac{r}{\mathrm{AU}}\right)^{-3/2}\mathrm{g}\,\mathrm{cm}^{-2},  
\end{align}
a mid-plane density distribution
\begin{align}
  \rho_0 = 2.7\times10^{-9}
  \hat{\Sigma}\left(\frac{r}{\mathrm{AU}}\right)^{-39/14}\mathrm{g}\,\mathrm{cm}^{-3},  
\end{align}
which implies $p=-39/14$, {\bf ASK: should there be $\hat{T}$ and $\mu$ dependence for midplane density?}
a temperature profile
\begin{align}\label{mmsn_temp}
  T = 120\hat{T}\left(\frac{r}{\mathrm{AU}}\right)^{-3/7} \mathrm{K},
\end{align}
which implies $q=-3/7$, 
and the fiducial disk aspect-ratio 
\begin{align}\label{mmsn_epsilon}
  \epsilon = 3.36\times10^{-2}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\left(\frac{r}{\mathrm{AU}}\right)^{2/7}.
\end{align}
%0.033576258
In the above expressions, $\hat{\Sigma}$ and
$\hat{T}$ are dimensionless coefficients used to scale the model
relative to the MMSN. Their nomial values are unity. In addition, we
use the opacity model {\bf(reference?)}
 \begin{align}
   \kappa_d &= 2 \hat{\kappa}_d \left(\frac{T}{100\mathrm{K}}\right)^2
   \mathrm{cm}^2\,\mathrm{g}
    =
   2.88\hat{\kappa}_d\hat{T}^2\left(\frac{r}{\mathrm{AU}}\right)^{-6/7}\mathrm{cm}^2\,\mathrm{g}^{-1},   
 \end{align}
where the second equality follows from the model temperature profile
above, and $\hat{\kappa}_d$ has similar meaning as $\hat{\Sigma}$ and
$\hat{T}$.

### 2015-01-26 application.tex
editing numerical results, redid mmsn application for 50au and corrected espilon to include mu dependence

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).  
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model, 
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{9.4\times10^{-3}}{(\gamma
    -1)}\zeta^{2/7},
\end{align}
and we recall $\beta < \beta_\mathrm{crit}$ permits the fundamental 
isothermal VSI to operate.

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the MMSN with $\mu = 2$ and
$\gamma=1.4$. Although it is possible to satisfy
$\beta<\beta_\mathrm{crit}$ at small radii by choosing a sufficiently
large $\khat$, our linear calculations suggest growth rates eventually
decay as $\khat\to\infty$. We thus restrict our attention to
$\khat\lesssim 10^2$.

Fig. \ref{mmsn_bcrit_bcool} suggests that the fundamental isothermal
VSI can operate in the outer parts of the MMSN at radii on the order
of tens of AU. However, the thermal relaxation time $\beta$ is not
much shorter than $\beta_\mathrm{crit}$, suggesting the VSI may be
weak.

@@

\subsection{Numerical example}
Consider perturbations with $\khat=10$ in the 
the MMSN at $r=30\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2$. Then we find
the dimensionless thermal relaxation time, using
Eq. \ref{tc_newton_cool}, is
\begin{align*}
  \beta \simeq 0.046. 
\end{align*}
At $r=30\mathrm{AU}$, the disk aspect ratio is $\epsilon \simeq
0.058$ which implies, using Eq. \ref{iso_vsi_cond}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.062. 
\end{align*}
Thus, the disk supports the fundamental isothermal VSI, but only
marginally. Using the above value of $\beta$, MMSN disk parameters
and $\khat=10$ in our linear code, we find a growth rate of
\begin{align*}
  \nu = 2.7\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time of $t_g\simeq 60$
orbits.

 %only find one unstable mode -> close to marginal

 %larger kx

>>>>

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).  
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model, 
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7},
\end{align}
and we recall $\beta < \beta_\mathrm{crit}$ permits the fundamental 
isothermal VSI to operate.

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the MMSN with $\mu = 2.33$ and
$\gamma=1.4$. Although it is possible to satisfy
$\beta<\beta_\mathrm{crit}$ at small radii by choosing a sufficiently
large $\khat$, our linear calculations suggest growth rates eventually
decay as $\khat\to\infty$. We thus restrict our attention to
$\khat\lesssim 10^2$.

Fig. \ref{mmsn_bcrit_bcool} suggests that the fundamental isothermal
VSI can operate in the outer parts of the MMSN at radii on the order
of tens of AU. However, the thermal relaxation time $\beta$ is not
much shorter than $\beta_\mathrm{crit}$, so the VSI may be
weak.

@@

\subsection{Numerical example}
Consider  the MMSN at $r=50\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2.33$. Eq. \ref{mmsn_epsilon} gives $\epsilon \simeq
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The disk supports the fundamental isothermal VSI, but only
marginally. Using the above value of $\beta$, MMSN disk parameters and
$\khat=10$ in our linear code, we find a 
growth rate of 
\begin{align*}
  \nu = 4.0\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time $\sim 40$
orbits. 
%This mode was found to be the most unstable for
%$\khat=10$.

% and
% physically acceptable -> other modes
% are crazy

%We also solved the linear problem with varying $\khat$  
%and $\beta$ determined by Eq. \ref{beta_mmsn}. We find the most
%unstable wavenumber occurs at $\khat\simeq60$ with $\nu\simeq
%3.6\times10^{-3}\Omega_k$, i.e. only slightly faster than $\khat=20$.   
%However, the eigenfunctions for large $\khat$ were found to have large
%amplitude oscillations near the vertical boundaries. This is shown in
%Fig. \ref{mmsn_eigenW}. These modes may not be important since they
%mostly affect the disk atmosphere where there is little mass.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/eigenvectorW_mmsnkx10.ps}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0cm]{figures/eigenvectorW_mmsnkx60.ps}
  \caption{Most unstable VSI modes in the MMSN at
    $r=30\mathrm{AU}$. The eigenfunction $W$ for $\khat=20$ (top) and
    $\khat=60$ (bottom) are shown. The real (imaginary) part of $W$
    is plotted as the solid (dotted) lines.\label{mmsn_eigenW}}  
\end{figure}

% only find one unstable mode -> close to marginal 
% larger kx

### 2015-01-26 linear_adia.tex
adding youdin notes
almost done with numerical results section
editing numerical results, redid mmsn application for 50au and corrected espilon to include mu dependence

% \subsubsection{Qualitative condition for buoyancy to be negligible in
%   the presence of thermal relaxation}
% {\bf this section might be wrong, but keep for reference for now}
% We have shown that when $\gamma>1$ so that there is a positive
% vertical entropy gradient and perturbations evolve
% adiabatically ($t_c\to\infty$) that the VSI is strongly
% surpressed. This is the limit where the term proportional to
% $t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
% becomes negligible compared to the term proportional to
% $\gamma/\Gamma-1$. The latter term is associated with the vertical
% entropy gradient, or buoyancy frequency squared, $N_z^2$.

% Reversing the argument, we expect to recover the VSI when the bouyancy
% frequency becomes negligible compared to the cooling rate, or 
% \begin{align}
%   t_c \ll \frac{\Omega}{\gamma N_z^2}. 
% \end{align} 
% Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
% for $N_z^2$ appropriate for vertically isothermal disks, we expect
% \begin{align}
%   \beta \ll
%   \frac{1}{H^2_\mathrm{iso}\left(\gamma-1\right)\left(\frac{\p\ln\rho}{\p
%           z}\right)^2} 
% \end{align}
% is required to enable the VSI, where we used $\Omega\sim\Omega_k$.  
% This inequality is satisfied at the
% midplane and as $|z|\to\infty$. It can be satisifed at all heights if
% it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
% Using the full expression for the density field for vertically isothermal
% disks (Eq. \ref{vertiso_eqm}), we expect 
% \begin{align}
%   \beta \ll \frac{27}{4(\gamma-1)}\epsilon^2,
% \end{align}
% i.e. $\beta\ll O(\epsilon^2)$, is a sufficient condition for
% buoyancy to be negligible.

@@

\subsubsection{The effect of introducing a small but finite
  thermal relaxation time} 
It is instructive to first ask  the more analytically tractable
question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c$ from zero to a small but finite value? For sufficiently small
$\beta$, for which the perturbations may be considered 
nearly-isothermal, we expect the new solutions to only differ slightly
from the explicit isothermal solutions developed in \S\ref{iso_poly}.

@@

At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} in
the limit $\khat\to\infty$ we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
We seek solutions with
\begin{align}
  \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \khat\hat{\omega} \text{ finite}.
\end{align}
Then
\begin{align}
  \khat\hat{\omega} = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2}. 
\end{align}
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental isothermal VSI to operate.  For
$\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$ we find
$\beta_\mathrm{crit}=0.125$, in agreement with Fig. \ref{relax_bcrit}
at large $\khat$.

@@

% The pair of ODEs for $W$ and $Q$
% (Eq. \ref{ode_w}---\ref{ode_Q}) with these approximations and 
% written with dimensionless variables, are
% \begin{align}
%   &\hat{\sigma}^2\left(Q +  \hat{k}^2W\right) \notag\\ 
%   &= \hat{z} \left(1 +
%     \ii\epsilon q \hat{k} \right)\left[W^\prime - \zhat \left(W -
%       Q\right)\right] - \left[W^\prime - \zhat \left(W -
%       Q\right)\right]^\prime\label{nearly_iso1},\\
% &\beta \hat{\sigma}^2\left(W - \gamma Q\right) + \ii\hat{\sigma}
%     \left(W-Q\right)\notag\\
%    & = -\beta\zhat\left(\gamma-1\right) \left[W^\prime - \zhat \left(W -
%       Q\right)\right]\label{nearly_iso2},
% \end{align}
% where we recall $\beta = t_c\Omega_k$ is the dimensionless cooling
% time. It is evident that for $\beta\equiv 0$, we recover
% Eq. \ref{iso_ode3}. In addition, from the expression of the vertical
% velocity perturbation (Eq. \ref{lin_vz}) we have
% \begin{align}
%   W^\prime - \zhat \left(W -
%       Q\right) = \ii\hat{\sigma}c_s\delta v_z. \label{nearly_vz}
% \end{align}

% %is valid for
% %non-isothermal perturbations $\gamma>1$ if the thermal relaxation
% %(cooling) timescale $t_c\equiv0$.

% % Then we may linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2}

% To determine the (assumed) small effect of introducing a short thermal
% relaxation, we linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2} about a known
% solution for $\beta\equiv0$. Let 
% \begin{align}\label{nearly_iso_pert}
%   \beta \to 0 + \delta\beta,\, W\to W+\delta W,\, Q \to Q+\delta
%   Q,\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
% \end{align} 
% where $W=Q$ is the polynomial solution described in \S\ref{iso_poly}
% with $\hat{\sigma}^2$ given by Eq. \ref{sig2_iso}. Inserting
% Eq. \ref{nearly_iso_pert} into
% Eq. \ref{nearly_iso1}---\ref{nearly_iso2} and keeping only first order
% terms, we obtain
% \begin{align}
%   &2\hat{\sigma}\delta\hat{\sigma}\left(1+\hat{k}^2\right)W +
%   \hat{\sigma}^2 \left(\delta Q + \hat{k}^2\delta W\right)\notag\\
%   &= \zhat \left(1+\ii\epsilon q \hat{k}\right)\left[\delta W^\prime -
%     \zhat \left(\delta W - 
%       \delta Q\right)\right] - \delta W^{\prime\prime}\notag\\
%   &\phantom{=} + \left[\zhat \left(\delta W -
%      \delta Q\right)\right]^\prime,\label{nearly_iso3}\\
% &\ii\hat{\sigma} \left(\delta W - \delta Q\right) =
% \delta\beta\left(\gamma-1\right)\left(\hat{\sigma}^2W - \hat{z}W^\prime\right).\label{nearly_iso4}
% \end{align}
% This is a pair of inhomogeneous ODEs with a forcing by $W$. The
% homogeneous problem is, of course, just the ODE for $\beta=0$ solved
% in \S\ref{iso_poly}. Hence we only seek the particular solution.

% We could proceed by considering the entire set of polynomial
% solutions for $W$, but it is simplest to look at the fundamental
% mode. Thus we consider the `basic state' 
% \begin{align*}
%   W = \hat{z},\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
%   \ii\epsilon q \hat{k},
% \end{align*}
% then
% \begin{align}
%   \delta W = b \zhat^3,
% \end{align}
% where $b$ is a constant to be determined. Eliminating $\delta Q$ from
% Eq. \ref{nearly_iso3}---\ref{nearly_iso4}, inserting the above
% expressions for $W$ and $\delta W$ and balancing coefficients of
% $\zhat$ and $\zhat^3$, we find
% \begin{align}
%  & 2\left(1+\ii\epsilon q \hat{k}\right) \delta\hat{\sigma} \notag\\& = -6b\hat{\sigma} -
% 2\ii \delta \beta\left(\gamma-1\right)\left(\hat{\sigma}^2-1\right) - \ii
%   \hat{\sigma}^2\delta\beta\left(\gamma-1\right),\\
% & 0=  2b\hat{\sigma}+\ii\delta\beta
% \left(\gamma-1\right)\left(\hat{\sigma}^2 - 1\right),
% \end{align}
% where the expression for $\hat{\sigma}^2$ was used.

>>>>

@@

\subsubsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
It is instructive to first ask  the more analytically tractable
question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c$ from zero to a small but finite value? For sufficiently small
$\beta$, for which the perturbations may be considered 
nearly-isothermal, we expect the new solutions to only differ slightly
from the explicit isothermal solutions developed in \S\ref{iso_poly}.

@@

At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} in
the limit $\khat\to\infty$ we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
We seek solutions with
\begin{align}
  \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \khat\hat{\omega} \text{ finite}.
\end{align}
Then
\begin{align}
  \khat\hat{\omega} = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2}. 
\end{align}
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental isothermal VSI to operate.  For
$\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$ we find
$\beta_\mathrm{crit}=0.125$, in agreement with Fig. \ref{relax_bcrit}
at large $\khat$.

@@
